---
title: "1_process_raw_data"
author: "Toon Van Daele"
date: "2024-07-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# Pre-processing raw data

This script does pre-processing and first checks on time series

It reads the datasets downloaded from multiple sources (filesnames 0_load_xxx.R).

After processing, the data are stored in a single data frame for further analysis

## Load raw datasets

```{r}
df_voss_p <- readRDS(file = "./data/interim/vosselaar_p.rds")
df_heren_me <- readRDS(file = "./data/interim/herentals_ME.rds")
df_kmi_retie <- readRDS(file = "./data/interim/kmi_retie.rds")
```

## Process Waterinfo data

### Vosselaar precipitation

Check quality code all '130' -> ok

```{r}
df_voss_p %>%
  filter(!`Quality Code` == "130")
```

add column with station name and variable

```{r}
df_voss_p %>%
  mutate(loc = "Vosselaar",
         date = as.Date(Timestamp),
         var = "P") %>%
  dplyr::select(loc, date, var, value = Value) -> df_voss

head(df_voss)
```

### Herental PET - Penman Monteith 

Check Quality code == 130 (ok)

```{r}
df_heren_me %>%
  filter(!`Quality Code` == "130") %>%
  group_by(`Quality Code`) %>%
  summarise(n = n())
```

(quality code table from https://www.milieuinfo.be/confluence/display/DDOV/Kwaliteitsvlaggen+hydrometrie-data+HIC+en+VMM

code '-1' = missing      - replaced with 0 for the moment (to be changed later better estimation possible )
code '180' = estimated   - we keep the estimated value

Check when the missing data occur most

```{r}
df_heren_me %>%
  filter(`Quality Code` == "-1") %>%
  mutate(year = year(Timestamp)) %>%
  group_by(year) %>%
  summarise(n = n())
```

Set NA values to 0 (better estimation possible later)

```{r}
df_tmp <- df_heren_me %>%
  mutate(Value = ifelse(is.na(Value), 0, Value))
```

```{r}
df_tmp %>%
  mutate(loc = "Herentals",
         date = as.Date(Timestamp),
         var = "PETpm") %>%
  dplyr::select(loc, date, var, value = Value) -> df_herentals

head(df_herentals)
```

```{r}
ggplot(df_herentals, aes(x = date, y = value)) + geom_point()
```


We have negative values!! It's a problem with the original data from waterinfo.

For the moment we set negative values equal to zero.
Check later if KMI PET data is better

```{r}
df_herentals %>%
  mutate(value = ifelse(value < 0, 0, value)) -> df_herentals
```


```{r}
ggplot(df_herentals, aes(x = date, y = value)) + geom_point()
```



## Process KMI data

```{r}
glimpse(df_kmi_retie, )
```

We are only interested in the precipitation and temperature data (for the moment)

NA values for precipitation mean no precipitation (change to 0)

```{r}
df_tmp <- df_kmi_retie %>%
  mutate(datetime = as.POSIXct(timestamp, format="%Y-%m-%dT%H:%M")) %>%
  dplyr::select(datetime, P = precip_quantity, t = temp) %>%
  replace_na(list(P = 0)) %>%
  arrange(datetime)

head(df_tmp)
```


Some visual checks

```{r}
df_tmp %>%
  ggplot(aes(x = datetime, y = t)) + geom_line()
```

The hourly data need to be summarised to daily data:
  - temperature -> mean()
  - precipitation -> sum()

```{r}
df_tmp %>%
  mutate(date = as.Date(datetime)) %>%
  group_by(date) %>%
  summarise(P = sum(P),
            t = mean(t)) -> df_tmp_daily
```

Long format

```{r}
df_tmp_daily %>%
  mutate(loc = "Retie") %>%
  pivot_longer(cols = c("P", "t"), names_to = "var") -> df_retie_long

head(df_retie_long)
```


## Join datasets

```{r}
df_voss %>%
  rbind(df_herentals, df_retie_long) %>%
  mutate(var = as.factor(var),
         loc = as.factor(loc)) -> df_meteo

unique(df_meteo$var)
```


## Some visual checks

```{r}
df_meteo %>%
  filter(var == "P") %>%
  ggplot(aes(x = date, y = value)) + geom_line()
```

```{r}
df_meteo %>%
  filter(loc == "Retie" & var == "t" | loc == "Herentals" & var == "PETpm") %>%
  ggplot(aes(x = date, y = value, colour = var)) + geom_line()
```

```{r}
df_meteo %>%
  filter(loc == "Retie" & var == "t" | loc == "Herentals" & var == "PETpm") %>%
  dplyr::select(loc, date, var, value) %>%
  pivot_wider(names_from = c("var", "loc"), values_from = "value") %>%
  ggplot(aes(x = t_Retie, y = PETpm_Herentals)) + geom_point()
```



